---
output:
  word_document: default
  html_document: default
---
If our molecular pathway disruption clusters represent distinct biological states
apart from tissue of origin, they may have different prognoses within cancer
types or across all cancers. These analyses are limited by confounding factors
of age and stage at diagnosis, sex, ethnicity, and tissue-specific disease
progression. To explore these ideas, we used Bayesian inference to test models
of survival using public longevity data from the CDC, accounting for age, gender
and ethnicity. Our model estimates the effect of cancer type and
cluster-specific cancer effects independently, resulting in a cancer and
cluster-specific estimate of the aging rate multiplier, k.

To understand the model, we show in **Figure 7A** simulated survival time curves
for four related groups. We start with a cohort of women aged 30 and show their
expected survival (light blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (yellow). Compare this
to a randomly selected group of women ranging in age from 30 to 70 years old
(dark blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (red). We can see that for different tissues, simply due to the change
in distribution of age-at-diagnosis, we should expect equally deadly cancers to
have *different* survival curves. The model takes this into account and the
aggressiveness of the cancer can be estimated accurately without the confounding
effect of age.

```{r, results='asis'}
cat("\\newpage")
```

#### Figure 7
```{r child='../code/survival/survfig.Rmd'}
```

**Figure 7: Comparison of different cancers, simulated and actual results
results.** To illustrate the risk model, we compare survival of four simulated
cohorts. A) A group consisting of randomly selected 30 year old
women (light blue) vs. an equal sized group of 30 year old women with cancer, k =
2 (light orange). To illustrate how the mix of ages alters the survival curve,
we compare a randomly selected group of people with ages uniformly distributed
from 30 to 70 (dark blue) and a randomly selected group with uniformly
distributed ages and cancer k = 2 (dark orange). B,C,D,E) We compare a randomly
selected group with correct age distribution for the given cancer (turqoise) to
the model predicted survival for the given average k (blue) and the actual
survival for the patients in our dataset (orange). As can be seen, Stomach
Cancer (STAD), Glioblastoma (GBM), and Pancreatic Cancer (PAAD) are all very
deadly. The modeled results fit the general trend for the actual patients.
Thyroid Cancer (THCA) on the other hand has very little if any effect on our
estimates of life expectancy, in fact there may be a slight benefit which may be
associated with a bias in diagnosis towards patients who are more health
conscious, or higher socioeconomic status, with more access to care compared to
random population members.

```{r, results='asis'}
cat("\\newpage")
```

We found that cancer types, as expected, have a range of prognoses relative to
the general population. For example in **Figure 7B-E** we see three particularly
deadly cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one
cancer where diagnosis apparently decreases risk relative to background
(Thyroid: THCA). Cancers with posterior probability for relative risk of less
than 1 such as THCA should be interpreted carefully. This Bayesian model is a
model for a state of information. The information that a person is diagnosed
with cancer may lead us to expect that they will live a shorter time than the
general population of matched age (k > 1), or a longer time than the general
population of matched age (k < 1). One mechanism for a shorter result is that
the cancer is aggressive and we can expect it to rapidly injure the body,
causing death. One explanation for longer results may be that the cancer is
relatively mild, and therefore diagnosis is potentially an informational signal
that the patient is health conscious, with the comparison group having more
people whose cancers go undiagnosed. It's important to note that the diagnosis
can increase our expectation of life relative to the comparison group, even if
it decreases the expectation of life of the individual relative to the
counterfactual where they did not have cancer.

The degree to which a given cancer accelerates mortality can be determined by
the $k_{tis}$ multiplier (**Figure 8**) multiplied by the cluster modifier
(**Figure 9**). Looking at the tissue specific multiplier, the least deadly was
prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
risk multiplier of between 15 and 22 relative to the background risk of death in
the population. This undoubtedly is influenced by the fact that relatively young
patients are affected by GBM and that it is extremely deadly. Among the
deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
in relative risk of age-adjusted death, the remaining cancer types ranged from
about 1 to 5 in magnitude. A rate of 1 represents patients who die relatively
close to the age predicted purely from their demographics independent of the
information that they have the given cancer.

```{r, results='asis'}
cat("\\newpage")
```

#### Figure 8
```{r fig.width=3, message=FALSE, warning=FALSE}
pcanc+mytheme
```

**Figure 8 Tissue specific effects** Effects are compared between cancer type by
showing the probability density for the posterior value of the $k_{tis}$
multiplier. These values reflect the general deadliness of the cancer type
across all clusters. GBM, PAAD, and SKCM stand out as having values in the 10 to
20 range with a wide degree of uncertainty. Other cancers such as Prostate
(PRAD) and Thyroid (THCA) show values down near or even below 1 suggesting that
these populations may have healthier lifestyles and more access to care. Most
cancer types fall in the range 2-7 as might be expected.

```{r, results='asis'}
cat("\\newpage")
```

#### Figure 9

```{r child='../code/figure9.Rmd'}
```

**Figure 9 Cluster specific $k_{cl}$ values for each cancer type.** The overall
$k$ value is the product of $k_{tis} \times k_{cl}$. The cluster-specific value
represents the relative aggressiveness of each cluster within the cancer type.
For example clusters 1, 2, and 6 Glioblastoma (GBM) are apparently much less
deadly than other types of GBM. A similar situation can be seen with Pancreatic
Cancer (PAAD) where 1, 6, and 8 are less deadly compared to clusters 7, 9, 10.
For some cancer types there are few data points and little ability to estimate
values with precision (DLBC and KICH for example).


```{r, results='asis'}
cat("\\newpage")
```

Our estimates of tissue-specific cluster effects, in contrast, ranged from less
than 1 up to about 3 or 4, reflecting that some clusters are either less deadly
or more deadly than other clusters within each cancer type (**Figure 9**). A
cluster-specific rate of 1 represents typical rate for this tissue type. For
several cancers ( _e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
indistinguishable from the prior, reflecting that either there were too few
mortalities in the data to make an estimate (as expected for PRAD and THCA) or
two few samples, period. We did not observe cluster-specific trends that held true
across cancer types, which could result from different cancers having different
standards of care for example. In support of this interpretation, we also tested
a factored model which considered both cancer type and each cluster independent of
cancer type (not shown). Though we were able to successfully fit this model, it
is a special case of the more general model where cluster and tissue can
independently affect longevity, and there is no reason to believe that cluster
specific effects would necessarily remain constant across tissue types given how
widely the patients vary across tissue types as well as the different drug and
surgical treatments typically used for each tissue.
