---
output:
  word_document: default
  html_document: default
---
If our molecular pathway disruption classes represent distinct biological states
apart from tissue of origin, they may have different prognoses within cancer
types or across all cancers. Kaplan-Meier plots reveal large differences between
cancer types as expected, and some modest differences between our classes (not
shown). These analyses are profoundly limited by confounding factors of age and
stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
name the obvious ones. To explore these ideas, we tested models of survival
using public longevity data from the CDC, accounting for age, gender and
ethnicity. Our model assumes that at birth each person has a small initial
probability per unit time to die, and that this probability grows exponentially
in time at some baseline rate that matches the observed CDC data. The rate of
growth in risk is assumed to jump to a higher level at diagnosis, which we take
as a proxy for onset. This represents an accelerated aging type model. To fit
the model, we constrain the baseline risks using the available CDC life tables.
We then constrain the cancer specific multiplier using the longevity data within
the cancer tumor dataset. Our model estimates the effect of cancer type and
class-specific cancer effects independently, resulting in a cancer and class
specific estimate of the aging rate multiplier.

To understand the model, we show in **Figure 7A** simulated survival time curves
for four related groups. We start with a cohort of women aged 30 and show their
expected survival (light blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (yellow). Compare this
to a randomly selected group of women ranging in age from 30 to 70 years old
(dark blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (red). We can see that for different tissues, simply due to the change
in distribution of age-at-diagnosis, we should expect equally deadly cancers to
have *different* survival curves. The model takes this into account and the
aggressiveness of the cancer can be estimated without the confounding aspect of
age.

```{r, results='asis'}
cat("\\newpage")
```

#### Figure 7
```{r child='../code/survival/survfig.Rmd'}
```

**Figure 7: Comparison of different cancers, simulated and actual results
results.** A) To understand the risk model, we plot survival of four simulated
cohorts. Compare survival of a group consisting of randomly selected 30 year old
women (light blue) to an equal sized group of 30 year old women with cancer, k =
2 (light orange). To understand how the mix of ages alters the survival curve,
we compare a randomly selected group of people with ages uniformly distributed
from 30 to 70 (dark blue) and a randomly selected group with uniformly
distributed ages and cancer k = 2 (dark orange). B,C,D,E) We compare a randomly
selected group with correct age distribution for the given cancer (blue) to the
model predicted survival for the given average k (red) and the actual survival
for the patients in our dataset (CHOOSE COLOR HERE, NOT GREEN). As can be seen,
Stomach Cancer (STAD), Glioblastoma (GBM), and Pancreatic Cancer (PAAD) are all
very deadly. The modeled results fit the general trend for the actual patients.
Thyroid Cancer (THCA) on the other hand has very little if any effect on life
expectancy, in fact there may be a slight benefit which may be associated with a
bias in diagnosis towards patients who are more health conscious, or higher
socioeconomic status, with more access to care compared to random population
members.

```{r, results='asis'}
cat("\\newpage")
```

We found that cancer types, as expected, have a range of prognoses relative to
the general population. For example in **Figure 7B-E** we see three particularly
deadly cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one
cancer where diagnosis apparently decreases risk relative to background
(Thyroid: THCA). Cancers with posterior probability for relative risk of less
than 1 such as THCA may be those where diagnosis is associated with patients who
have relatively healthy behaviors, are seen regularly by doctors, and comply
with health recommendations.

The degree to which a given cancer accelerates aging can be determined by the
$k_{tis}$ multiplier (**Figure 8**) multiplied by the class specific modifier
(**Figure 9**). Looking at the tissue specific multiplier, the least deadly was
prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
risk multiplier of between 15 and 22 relative to the background risk of death in
the population. This undoubtedly is influenced by the fact that relatively young
patients are affected by GBM and that it is extremely deadly. Among the
deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
in relative risk of age-adjusted death, the remaining cancer types ranged from
about 1 to 5 in magnitude. A rate of 1 represents patients who die relatively
close to the age predicted purely from their demographics independent of the
information that they have the given cancer.

```{r, results='asis'}
cat("\\newpage")
```

#### Figure 8
```{r fig.width=3, message=FALSE, warning=FALSE}
pcanc+mytheme
```

**Figure 8 Tissue specific effects** Effects are compared between cancer type by
showing the probability density for the posterior value of the $k_{tis}$
multiplier. These values reflect the general deadliness of the cancer type
across all clusters. GBM, PAAD, and SKCM stand out as having values in the 10 to
20 range with a wide degree of uncertainty. Other cancers such as Prostate
(PRAD) and Thyroid (THCA) show values down near or even below 1 suggesting that
these populations may have healthier lifestyles and more access to care. Most
cancer types fall in the range 2-7 as might be expected.

```{r, results='asis'}
cat("\\newpage")
```

#### Figure 9

```{r child='../code/figure9.Rmd'}
```

**Figure 9 Class specific $k_{class}$ values shown for each cancer type.** The
overall $k$ value is the product of $k_{tis} \times k_{class}$. The class
specific value shows the relative aggressiveness of that particular class within
the tissue. For example class 1, 2, and 6 Glioblastoma (GBM) are apparently much
less deadly than other types of GBM. A similar situation can be seen with
Pancreatic Cancer (PAAD) where 1, 6, and 8 are less deadly compared to classes
7, 9, 10. For some cancer types there are few data points and little ability to
estimate values with precision (DLBC and KICH for example).


```{r, results='asis'}
cat("\\newpage")
```

Our estimates of tissue-specific class effects, in contrast, ranged from less
than 1 up to about 3 or 4, reflecting that some classes are either less deadly
or more deadly than other classes within each cancer types (**Figure 9**). A
class specific rate of 1 represents typical rate for this tissue type. For
several cancers ( _e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
indistinguishable from the prior, reflecting that either there were too few
mortalities in the data to make an estimate (as expected for PRAD and THCA) or
two few samples, period. We did not observe class-specific trends that held true
across cancer types, which could result from different cancers having different
standards of care for example. In support of this interpretation, we also tested
a factored model which considered both cancer type and each class independent of
cancer type (not shown). Though we were able to successfully fit this model, it
is a special case of the more general model where class and tissue can
independently affect longevity, and there is no reason to believe that class
specific effects would necessarily remain constant across tissue types given how
widely the patients vary across tissue types as well as the different drug and
surgical treatments typically used for each tissue.
